Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-12 doc/manuscript/translocation.qmd
Make three panel model figure and update main text

One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012]. Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amid this global loss of biodiverity, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016].
Although we know relatively little about the underlying mechanisms [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see \[@medzhitov2012\] for definitions). However, linking these evolutionary responses to population recoveries - as necessary for evolutionary rescue - is difficult and few clear examples exist. In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence \[e.g., @mendelson2019; @brannelly2021\].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

@@

Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).

@@

![Predicted growth and persistence of translocated populations. Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability \$\sigma\_{A_R}\$ and probability of recruitment $\omega$, given the parameterized, deterministic model. Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line). The red line shows where $\lambda = 1$. (B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$. The dashed line shows where $\omega = 0.3$, for comparison to (A). There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505.](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="450"}

![<!--# Evidence sentence -->Sensitivity analysis and empirical validation of the stage-structured MYL frog model. (A) Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch). Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3. (B) 100 simulated trajectories (black lines) that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}

@@

Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations.

@@

>>>>

One consequence of human alteration of the biosphere is an increase in emerging infectious diseases [@jones2008; @fisher2012].
Such diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and extinctions in a wide range of taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Amid this global loss of biodiverity, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016].
Although we know relatively little about the underlying mechanisms [@brannelly2021; @russell2020], host evolution is one possibility [@carlson2014; @searle2020].
This process is known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often results in increased host resistance and/or tolerance (see [@medzhitov2012] for definitions).
However, linking these evolutionary responses to population recoveries - as necessary for evolutionary rescue - is difficult and few clear examples exist.
In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in such populations could potentially be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019; @brannelly2021].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

@@

Even when incorporating demographic stochasticity and environmental stochasticity in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental variability in the presence of Bd), populations with high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50-year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability C; population 70550, surveyed for 16 years).

@@

![Predicted and observed growth and persistence of translocated populations. Population viability analysis of translocated MYL frogs quantified the long-run growth rate and extinction probability of frog populations: (A) Predicted long-run growth rate $\lambda$ for different values of yearly adult survival probability $\sigma_{A_R}$ and probability of recruitment $\omega$, given the parameterized, deterministic model. Colored points show the predicted $\lambda$ values for the twelve translocated populations when probability of recruitment into the adult class is $\omega = 0.3$ (indicated by the dashed line). The red line shows where $\lambda = 1$. (B) Predicted 50-year extinction probabilities of the 12 translocated populations given demographic stochasticity, environmental variability in $\omega$, and different mean values of $\omega$. The dashed line shows where $\omega = 0.3$, for comparison to (A). There are 6 lines at extinction probability = 1, 5 of which (70414--70619) are hidden beneath the line for 70505. (C) 100 simulated trajectories (black lines) from the population viability model that most closely matched the observed abundance trajectory of adult amphibians at site 70550 (light orange).](figures/pop_viability_figures_for_manuscript.jpg){#fig-viability width="450"}

@@

Second, we performed a local elasticity analysis on $\lambda$ with respect to parameters $\omega$, $\sigma_{A_R}$, and $F$ to determine how small changes in these parameters could influence the long-run deterministic growth rate of populations (@fig-viability-supp).

@@

![Sensitivity analysis of the stage-structured MYL frog model. Elasticity of $\lambda$ with changes in 3 parameters: $\sigma_{A_R}$ (yearly survival probability of naturally recruited adults), $\omega$ (yearly probability of successful recruitment from juveniles to adults), and *F (*number of eggs produced by a female frog in a year that successfully hatch). Elasticity is calculated at the default parameter values for each population and $\omega$ = 0.3.](figures/pop_viability_figures_for_supp.jpg){#fig-viability-supp width="450"}
